Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
section three (existing software) tidied up

Methods for exploring, understanding, and imputing missing data are more accessible now than they have ever been. In particular, the number of methods and implementations of imputation are seeing continued growth. Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets. This section discusses existing software for single and multiple imputation, and exploration.

@@

Each imputation method provides practical methods for different usecases, but most have different output structures, and do not have consistent interfaces in their implementation. This makes them inherently messy and challenging to integrate into an analysis pipeline. For example, combining different imputation methods from different pieces of software is not currently straightforward. \pkg{simputation} is different from other imputation packages, and resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returning a dataframe containing imputed values. This reduces the friction of working with other tools, but comes at the cost of identifying imputed values. An ideal approach would use consistent, simple data structures that work with other analysis tools, and help track missing values. This would make imputation outputs tidy, streamlining subsequent analysis.

@@

The MANET (Missings Are Now Equally Treated) software @Unwin1996 focussed on multivariate categorical data, with missingness explicitly added as a category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a Graphhical User Interface (GUI) for exploring missing data structure both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

@@

\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, data of school grades and test scores, a boxplot visualization with school grade and test score mapped to the x- and y-axis (Figure \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed (Figure \@ref(fig:gg-box-na)).

>>>>

Methods for exploring, understanding, and imputing missing data are more accessible now than they have ever been. Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets. This section discusses existing software for single and multiple imputation, and missing data exploration.

@@

Each imputation method provides practical methods for different usecases, but most have different output structures, and do not have consistent interfaces in their implementation. This makes them inherently messy and challenging to integrate into an analysis pipeline. For example, combining different imputation methods from different pieces of software is not currently straightforward. \pkg{simputation} resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returns a dataframe of imputed values. This reduces the friction of working with other tools, but comes at the cost of identifying imputed values. An ideal approach would use consistent, simple data structures that work with other analysis tools, and help track missing values. This would make imputation outputs tidy, streamlining subsequent analysis.

@@

The MANET (Missings Are Now Equally Treated) software @Unwin1996 focussed on multivariate categorical data, with missingness explicitly added as a category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

@@

\pkg{ggplot2} removes missing values with a warning (Figure \@ref(fig:warning)), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic. This has some limitations, shown in Figure \@ref(fig:gg-box-na), a boxplot visualization of school grade and test score mapped to the x- and y-axis . If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed (Figure \@ref(fig:gg-box-na)).
